---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# rcol

<!-- badges: start -->
[![R-CMD-check](https://github.com/PaulESantos/rcol/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PaulESantos/rcol/actions/workflows/R-CMD-check.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

`rcol` is an R client for the [Catalogue of Life](https://www.catalogueoflife.org)
via the [ChecklistBank](https://www.checklistbank.org) API
(<https://api.checklistbank.org>). It allows you to verify and match scientific names, retrieve hierarchical classifications, look up datasets, taxa, synonyms, vernacular names, and geographic distributions, and run ChecklistBank parsers — against **any** public dataset in ChecklistBank or the latest Catalogue of Life release. It is modelled on the conventions of [`rgbif`](https://docs.ropensci.org/rgbif/) and [`taxadb`](https://docs.ropensci.org/taxadb/).

> **Note on Origin & Architectural Evolution**: This package builds upon the original [`rcol` project by Catalogue of Life](https://github.com/CatalogueOfLife/rcol) (<https://catalogueoflife.github.io/rcol/>). This version introduces substantial architectural enhancements to simplify scientific workflows, including:
> - **Direct Scientific Name Support**: Pass scientific names directly (e.g., `"Panthera leo"` or `"Werneria nubigena"`) to downstream functions without needing prior manual ID lookups.
> - **Vectorized Multi-Species Queries**: Seamlessly pass character vectors of multiple species (e.g., `c("Werneria nubigena", "Panthera leo")`) and receive clean, concatenated tidy data frames.
> - **Species Traceability**: Preserves species association (`scientific_name` column) across synonyms, vernacular/common names, and distribution records.
> - **Clean Name Verification**: `col_check_name()` returns an essential 9-column resolution summary showing accepted names, status, and IDs.
> - **Flexible Classification Extraction**: `col_classification()` supports both Long format and desanided Wide format (`wide = TRUE`).

There are two parallel families of functions:

- **`col_*()`** convenience functions that target the **latest extended Catalogue of Life release** (`"3LXR"`). On first use, the release key is pinned in session (`col_key()`), ensuring long-running jobs are never disrupted by mid-session release updates. Use `col_refresh()` to re-pin.
- **`clb_*()`** universal low-level functions that accept an explicit `dataset =` argument (e.g., `"COL25"`, `2099`) to query any dataset in ChecklistBank.

## Installation

Install the package from GitHub:

```r
# install.packages("pak")
pak::pak("PaulESantos/rcol")

# or
# install.packages("remotes")
remotes::install_github("PaulESantos/rcol")
```

Documentation is published at <https://paulesantos.github.io/rcol/>.

## Quick start

### 1. Name Verification & Matching

```{r}
library(rcol)

# High-level clean verification (9 columns) for 1 or multiple species
col_check_name(c("Werneria nubigena", "Panthera leo", "Schinus molle"))

# Raw matching against the latest Catalogue of Life release
col_match("Panthera leo")

# Disambiguate with rank and nomenclatural code
col_match("Abies alba", rank = "species", code = "botanical")

# See all candidate matches (homonyms), not just the best one
col_match_verbose("Oenanthe")

# Match many names at once in parallel
col_match_checklist(c("Panthera leo", "Bufo bufo", "Abies alba"))
```

To match against a specific dataset instead of the latest COL release, use the `clb_*()` form with a `dataset =` key or alias:

```{r}
clb_match("Felis catus", dataset = "COL25")          # 2025 annual release
clb_match("Macrocystis pyrifera", dataset = 2099)    # public dataset by key
```

### 2. Classification, Synonyms, Vernaculars & Distribution

All functions work seamlessly with direct scientific names, IDs, character vectors, or match data frames:

```{r}
# Hierarchical classification in Wide or Long format
col_classification("Schinus molle", wide = TRUE)

# Synonyms for 1 or multiple species
col_synonyms(c("Werneria nubigena", "Panthera leo"))

# Vernacular / common names with species association (scientific_name column)
col_vernacular(c("Werneria nubigena", "Panthera leo"))

# Geographic distribution in tidy format
col_distribution(c("Werneria nubigena", "Schinus molle"))

# Details of a name usage
col_usage("Panthera leo")
```

### 3. Tree Navigation & Search

```{r}
# Walk the COL tree roots (Domains / Realms)
roots <- col_tree()

# Get direct child species under a genus
col_children("Panthera")

# Full-text search with rank and status filters
col_usage_search("Panthera", rank = "species", status = "accepted")
```

### 4. Parsers

```{r}
# Parse a scientific name into its atomic components
clb_parse_name("Abies alba Mill. var. alpina")
```

## Catalogue of Life Releases

The Catalogue of Life is published in two cadences (monthly and annual) and two flavours (base and extended `XR`). Every `clb_*()` function takes a `dataset =` argument; helpful aliases resolve to the latest of each:

```{r}
clb_col_release("monthly")                  # "3LR"   latest monthly base
clb_col_release("monthly", extended = TRUE) # "3LXR"  latest monthly extended
clb_col_release("annual")                   # e.g. "COL25"

clb_usage_search("Felidae", dataset = "COL25")
```

## Configuration

The base URL defaults to the production API and can be redirected with the `CLB_BASE_URL` environment variable:

```r
Sys.setenv(CLB_BASE_URL = "https://api.dev.checklistbank.org")
```

## Citation

To cite `rcol` in publications, please use:

```r
citation("rcol")
```

- **Package Citation**: Santos Andrade, P. E. & Döring, M. (2026). *rcol: R Client for the Catalogue of Life / ChecklistBank API*. R package version 0.1.0. URL: https://github.com/PaulESantos/rcol
- **Catalogue of Life Data Citation**: Bánki, O., Döring, M., Ower, G., et al. (2026). *Catalogue of Life ChecklistBank API*. https://doi.org/10.48580/d4tm

## License

- **R Code**: Released under the [MIT License](https://opensource.org/licenses/MIT) © 2026 Paul Efren Santos Andrade, Markus Döring & Catalogue of Life.
- **Catalogue of Life Data**: Distributed under the [Creative Commons Attribution 4.0 International License (CC BY 4.0)](https://creativecommons.org/licenses/by/4.0/).
